A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data

### 2024-09-22 manuscript/manuscript.Rmd
update manuscript

In this second tutorial we illustrate how to fit a Bayesian hazard regression model to the masked response priming data set used in the first tutorial. Fitting (Bayesian or non-Bayesian) regression models to the data is important when you want to study how the shape of the hazard function depends on various predictors [@singerAppliedLongitudinalData2003].

There are several analytic decisions one has to make. First, one has to select an analysis time window, i.e., a contiguous set of bins for which there is enough data for each participant. Second, given that the dependent variable is binary, one has to select a link function (see Figure 3).

@@

In the case of a large-*N* design without repeated measurements, the parameters of a discrete-time hazard model can be estimated using standard logistic regression software (after expanding the typical person-trial-oriented data set into a person-trial-bin-oriented data set (see Tutorial 1); @allisonSurvivalAnalysisUsing2010). When there is clustering in the data, as in the case of a small-*N* design with repeated measurements, the parameters of a discrete-time hazard model can be estimated using population-averaged methods (e.g., Generalized Estimating Equations), and Bayesian or frequentist generalized linear mixed models [@allisonSurvivalAnalysisUsing2010].

@@

To get the posterior distribution of the parameters given the data, we need to specify the prior distribution of the parameters. The middle column of Figure 4 shows seven examples of prior distributions on the logit and/or cloglog scales.

@@

To gain a sense of what prior *cloglog* values would approximate a uniform distribution on the hazard probability scale, we followed Kurz's approach and simulated a large number of draws from the Uniform(0,1) distribution, converted them to the cloglog metric, and fitted a skew-normal model (due to the asymmetry of the cloglog link function). Row E shows that using a skew-normal distribution with a mean of -0.59, a standard deviation of 1.26, and a skewness of -4.22 as a prior on the cloglog scale, approximates a uniform distribution on the probability scale. 
However, because hazard values below .5 are more likely in RT studies, using a skew-normal distribution with a mean of -1, a standard deviation of 1, and a skewness of -2 as a prior on the cloglog scale (row F), might be a good weakly informative prior for the alpha parameters or intercept(s) in a cloglog-hazard model. 
A skew-normal distribution with a mean of -0.2, a standard deviation of 0.71, and a skewness of -2.2 might be a good weakly informative prior for the beta parameters in a cloglog-hazard model as it gently regularizes p towards .6321 (i.e., a zero effect on the cloglog scale).

@@

\scriptsize
```{r data-M1, echo=T}
# remove unnecessary columns before fitting a model
M1_data <- ptb_data %>% select(-c(bl,tr,trial,period, period_9,d9))

@@

Estimating model_M1 took about 70 minutes on a MacBook Pro (Sonoma 14.6.1 OS, 18GB Memory, M3 Pro Chip).

@@

For the second model, we use a third-order polynomial specification of TIME for the baseline condition (blank prime), and assume that the effects of prime-target congruency and trial number are proportional and additive, and that the effect of trial number is linear.

@@

@@

Based on previous work [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhenDoesInhibition2022] and because cognition is likely the behavior of a non-linear dynamical system [ref], we relax all three assumptions in model 4.

@@

Use WAIC to compare models (ref McElreath, Kurz)

## Plot congruency effects and subject-specific fits for the selected model

@@

@@

@@

@@

In this final tutorial we illustrate how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of @panisWhatShapingRT2016, which includes two independent variables: mask type and prime type.

@@

Finally, as explained by @kelsoOutlineGeneralTheory2013, it is crucial to first have a precise description of the macroscopic behavior of a system (here: h(t) and ca(t) functions) in order to know what to derive on the microscopic level. EHA can thus solve the problem of model mimicry, i.e., the fact that different computational models predict the same mean RTs as observed in the empirical data, but not necessarily the same detailed shape of the empirical RT distributions. Also, fitting parametric functions or computational models to data without studying the shape of the empirical discrete-time h(t) and ca(t) functions can miss important features in the data [@panisStudyingDynamicsVisual2020; @panisWhatShapingRT2016].

>>>>

In this second tutorial we illustrate how to fit Bayesian hazard regression models to the masked response priming data set used in the first tutorial. Fitting (Bayesian or non-Bayesian) regression models to the data is important when you want to study how the shape of the hazard function depends on various predictors [@singerAppliedLongitudinalData2003].

There are several analytic decisions one has to make when fitting a hazard model. First, one has to select an analysis time window, i.e., a contiguous set of bins for which there is enough data for each participant. Second, given that the dependent variable is binary, one has to select a link function (see Figure 3).

@@

In the case of a large-*N* design without repeated measurements, the parameters of a discrete-time hazard model can be estimated using standard logistic regression software (after expanding the typical person-trial-oriented data set into a person-trial-bin-oriented data set; @allisonSurvivalAnalysisUsing2010). When there is clustering in the data, as in the case of a small-*N* design with repeated measurements, the parameters of a discrete-time hazard model can be estimated using population-averaged methods (e.g., Generalized Estimating Equations), and Bayesian or frequentist generalized linear mixed models [@allisonSurvivalAnalysisUsing2010].

@@

To get the posterior distribution of each parameters given the data, we need to specify the prior distribution of each parameter. The middle column of Figure 4 shows seven examples of prior distributions on the logit and/or cloglog scales.

@@

To gain a sense of what prior *cloglog* values would approximate a uniform distribution on the hazard probability scale, we followed Kurz's approach and simulated a large number of draws from the Uniform(0,1) distribution, converted them to the cloglog metric, and fitted a skew-normal model (due to the asymmetry of the cloglog link function). Row E shows that using a skew-normal distribution with a mean of -0.59, a standard deviation of 1.26, and a skewness of -4.22 as a prior on the cloglog scale, approximates a uniform distribution on the probability scale. 
However, because hazard values below .5 are more likely in RT studies, using a skew-normal distribution with a mean of -1, a standard deviation of 1, and a skewness of -2 as a prior on the cloglog scale (row F), might be a good weakly informative prior for the intercept(s) in a cloglog-hazard model. 
A skew-normal distribution with a mean of -0.2, a standard deviation of 0.71, and a skewness of -2.2 might be a good weakly informative prior for the beta parameters in a cloglog-hazard model as it gently regularizes p towards .6321 (i.e., a zero effect on the cloglog scale).

@@

\scriptsize
```{r data-priors-M1, echo=T}
# remove unnecessary columns before fitting a model
M1_data <- ptb_data %>% select(-c(bl,tr,trial,period, period_9,d9))

@@

Estimating model M1 took about 70 minutes on a MacBook Pro (Sonoma 14.6.1 OS, 18GB Memory, M3 Pro Chip).

@@

For the second model, we use a third-order polynomial specification of TIME for the baseline condition (blank prime), and again assume that the effects of prime-target congruency and trial number are proportional and additive, and that the effect of trial number is linear. We first remove unncessary columns and specify our priors.

\scriptsize
```{r data-priors-M2, echo=T}
# remove unnecessary columns
M2_data <- ptb_data %>% select(-c(bl,tr,trial,period, d6, d7, d8, d9, d10, d11, d12, d13, d14, d15))

# Specify priors
priors_M2 <- c(
  set_prior("skew_normal(-0.2,0.71,-2.2)", class = "b"),      
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),    
  set_prior("normal(0, 1)", class = "sd"),            
  set_prior("lkj(2)", class = "cor")                 
)
```
\normalsize

Now we can fit model 2.

\scriptsize
```{r fit-model-M2, eval=F, echo=T}
plan(multicore)

model_M2 <-
   brm(data = M2_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + period_9 + I(period_9^2) + I(period_9^3) + 
                          condition + trial_c +
                          (1 + period_9 + I(period_9^2) + I(period_9^3) + 
                          condition + trial_c | pid),
       prior = priors_M2,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M2")
```
\normalsize

Estimating model M2 took about 144 minutes.

@@

For the third model, we use a third-order polynomial specification of TIME for the baseline condition (blank prime), and relax the proportionality assumption for the predictor variables congruency (variable "condition") and trial number (variable "trial_c"). We use the same data set and priors as for model 2.

\scriptsize
```{r fit-model-M3, eval=F, echo=T}
M3_data <- M2_data
priors_M3 <- priors_M2
plan(multicore)

model_M3 <- 
   brm(data = M3_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + # Note that duplicate terms in the model formula are ignored
                           condition*period_9 + 
                           condition*I(period_9^2) + 
                           condition*I(period_9^3) +
                           trial_c*period_9 + 
                           trial_c*I(period_9^2) + 
                           trial_c*I(period_9^3) +
                           (1 + condition*period_9 +
                           condition*I(period_9^2) +
                           condition*I(period_9^3) +
                           trial_c*period_9 + 
                           trial_c*I(period_9^2) + 
                           trial_c*I(period_9^3) | pid),
       prior = priors_M3,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M3")
```
\normalsize

Estimating model M3 took about 268 minutes.

@@

Based on previous work [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhenDoesInhibition2022] and because cognition is likely the behavior of a non-linear dynamical system [ref], we relax all three assumptions in model 4. We use the same data set and priors as for model 2.

\scriptsize 
```{r fit-model-M4, eval=F, echo=T}
M4_data <- M2_data
priors_M4 <- priors_M2
plan(multicore)

model_M4 <- 
   brm(data = M4_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + # Note that duplicate terms in the model formula are ignored
                          condition*period_9*trial_c + 
                          condition*period_9*I(trial_c^2) + 
                          condition*I(period_9^2)*trial_c +
                          condition*I(period_9^2)*I(trial_c^2) +
                          condition*I(period_9^3) +
                          trial_c*I(period_9^3) +
                          (1 +  condition*period_9*trial_c +
                                condition*period_9*I(trial_c^2) + 
                                condition*I(period_9^2)*trial_c +
                                condition*I(period_9^2)*I(trial_c^2)  
                                condition*I(period_9^3) +
                                trial_c*I(period_9^3) | pid),
       prior = priors_M4,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M4")
```
\normalsize

Estimating model M4 took about 8 hours.

@@

We can compare the four models using the Widely Applicable Information Criterion (WAIC) and Leave-One-Out (LOO) cross-validation, and look at model weights [@mcelreathStatisticalRethinkingBayesian2018; @kurzAppliedLongitudinalDataAnalysis2023].

\scriptsize 
```{r compare-models, echo=T}
model_M1 <- readRDS("../Tutorial_2_Bayesian/models/model_M1.rds")
model_M2 <- readRDS("../Tutorial_2_Bayesian/models/model_M2.rds")
model_M3 <- readRDS("../Tutorial_2_Bayesian/models/model_M3.rds")
model_M4 <- readRDS("../Tutorial_2_Bayesian/models/model_M4.rds")

#model_M1  <- add_criterion(model_M1, c("loo", "waic"))
#model_M2  <- add_criterion(model_M2, c("loo", "waic"))
#model_M3  <- add_criterion(model_M3, c("loo", "waic"))
#model_M4  <- add_criterion(model_M4, c("loo", "waic"))

loo_compare(model_M1, model_M2, model_M3, model_M4, criterion = "loo") %>% print(simplify = F)
loo_compare(model_M1, model_M2, model_M3, model_M4, criterion = "waic") %>% print(simplify = F)

# model weights
model_weights(model_M1, model_M2, model_M3, model_M4, weights = "loo") %>% round(digits = 3)

model_weights(model_M1, model_M2, model_M3, model_M4, weights = "waic") %>% round(digits = 3)
```
\normalsize

Clearly, both weighting schemes prefer model M4.

## Plot congruency effects and subject-specific fits for model M4.

@@

Figure 5

@@

Figure 6

@@

model and comparison and note on fitting

Figure 7

@@

In this final tutorial we illustrate how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of @panisWhatShapingRT2016, which includes two independent variables: mask type and prime type. As we use the same functional programming approach as in tutorial 1, we simply present the sample-based functions for participant 6 in Figure 8. Note the negative compatibility effect in the hazard and conditional accuracy functions when a (relevant, irrelevant, or lines) mask is present.

(ref:descr-fig8-caption) Estimated discrete-time hazard, survivor, and conditional accuracy functions for participant 6, as a function of prime type (blank, congruent, incongruent) and mask type (no mask, relevant, irrelevant, lines).

```{r plot-2ivs, fig.cap = "(ref:descr-fig8-caption)", out.width='95%'}
knitr::include_graphics("../Tutorial_4_descriptive_2IV/Plot_2IV_for_subject6_PanisSchmidt.png")
```

@@

Finally, as explained by @kelsoOutlineGeneralTheory2013, it is crucial to first have a precise description of the macroscopic behavior of a system (here: h(t) and ca(t) functions) in order to know what to derive on the microscopic level. EHA can thus solve the problem of model mimicry, i.e., the fact that different computational models predict the same mean RTs as observed in the empirical data, but not necessarily the detailed shapes of the empirical RT hazard distributions. Also, fitting parametric functions or computational models to data without studying the shape of the empirical discrete-time h(t) and ca(t) functions can miss important features in the data [@panisStudyingDynamicsVisual2020; @panisWhatShapingRT2016].
